---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The chunk below sets the working directory, reads the data into an object called "activ" and then creates a data frame 'totsteps' containing the sum total of all steps taken per day, dropping missing cases.
```{r}
setwd("~/Dropbox/In process/MyRWork")
activ <- read.csv("activity.csv", header = TRUE)
# Need to express per day
library(dplyr)
byday <- group_by(activ, date)
totsteps <- summarize(byday,tot_steps=sum(steps, na.rm=TRUE))
```
## What is mean total number of steps taken per day?
In this section, we first create a histogram of the number of steps per day, and then go on to compute and report the mean and median number of steps per day.
``` {r}
hist(totsteps$tot_steps, main="Total Steps per Day")
mean_tot <- summarize(totsteps, m = mean(tot_steps))
med_tot <- median(totsteps$tot_steps)
```
The mean number of steps per day is `r mean_tot` and the median is `r med_tot`

## What is the average daily activity pattern?
To identify the daily pattern, it is first necessary to summarize the mean number of steps per 5-minute interval. I did this again using dplyr and omitting missing data.
Once the data were prepared, it's a small matter to create a time series plot. Note that this graph uses a data frame called 'mean_step' which contains the mean number of steps per 5-minute interval for the 288 intervals.  
``` {r}
by_int <- group_by(activ, interval)
mean_step <- summarize(by_int, mean_steps=mean(steps, na.rm=TRUE))

plot(mean_step$interval, mean_step$mean_steps, type="l", 
     main="Mean Steps per 5-minute Interval", xlab="Interval", 
     ylab="Average # Steps")

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
i <-which.max(mean_step$mean_steps)
maxint <- mean_step$interval[i]

```
From the plot we can see that `r maxint` is the most active of the day, on average. 

## Imputing missing values and re-do first part of the analysis
As specified in the assignment, first I computed the number of missing values in the dataset, and then proceed to impute values for the missing cases. For this assignment, I decided to replace missing values with the mean number of steps for the corresponding 5-minute interval.

After a couple of unsuccessful approaches, I consulted the class 
discussion, and followed the recommendation of Community TA Sebastián Jojoa Cruz. Thanks Sebastián! 

Note one odd calculation in this code: the vector 'ints' is meant to point to the means within the data frame of intervals; hence I wanted ints to map an interval number to a *row numbers* in that data frame. Since the intervals are numbered 5, 10, 15 etc. we can map an interval number to a row number muliplying the interval number by .2 and adding 1.
``` {r}
#Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
nas <- sum(is.na(activ$steps))
activ$stepsnm <- activ$steps
ind <- is.na(activ$stepsnm) # vector of rows with missing observations
ints <- 1+.2* activ$interval[ind] # identify row of intervals for each missing obs
activ$stepsnm[ind] <- mean_step$mean_steps[ints] #swap NA with mean for interval 
```
In the original dataset, there were `r nas` missing cases for the Number of Steps variable.

Now, having replaced the missing observations with the means for each interval, we repeat the first portion of the analysis.
```{r}
byday2 <- group_by(activ, date)
totsteps2 <- summarize(byday2,tot_steps2=sum(stepsnm, na.rm=TRUE))
hist(totsteps2$tot_steps2, main="Total Steps per Day", 
        xlab="Steps including Imputed Values")
mean_tot2 <- summarize(totsteps2, m2 = mean(tot_steps2))
med2 <-median(totsteps2$tot_steps2)
```
Here we see that the histogram is slightly more symmetric and mound-shaped than the original histogram. 

The new mean is `r mean_tot2` compared with the earlier value of `r mean_tot`.

The new median is `r med2`, compared with the earlier value of `r med_tot`-- no change at all.

## Are there differences in activity patterns between weekdays and weekends?
For this final portion of the analysis, it was necessary to create a factor variable with two levels: "weekend" and "weekday", and then recreate the daily activity pattern graphs in two panels, separating weekends and weekdays.

I chose to use the lattice package to create the plot. 
```{r}
wend <- c("Saturday", "Sunday")

activ <- activ %>% 
        mutate(day = ifelse(weekdays(as.Date(activ$date)) %in% wend, 
                 "Weekend", "Weekday"))

activ$day <- as.factor(activ$day)


#Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute #interval (x-axis) and the average number of steps taken, averaged across all weekday # days or weekend days (y-axis). 
by_int2 <- group_by(activ, day, interval)
mean_step2 <- summarize(by_int2, msteps =mean(stepsnm, na.rm=TRUE))

library(lattice)
xyplot(msteps ~ interval | day, data=mean_step2, type = "l",
       main="Comparison of daily actiivty patterns on Weekends and Weekdays", 
       ylab="Mean # of Steps", layout = c (1,2))

```


We note the general similarity in shape, with one particular difference. The weekend pattern has an addition spike roughly between intervals 1100-1300 that does not appear in such a pronounced way on weekdays.
